Inferring the existence and direction of causal associations in the face of measurement error

### 2017-02-01 docs/manuscript.Rmd
updated

fig_nums(
	"dag_figure",
	caption="Gene expression levels (blue blocks) and DNA methylation levels (green triangles) may be correlated but the causal direction is unknown. If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are three competing models for these variables. The causal inference test (CIT) attempts to delineate between them. a) Methylation causes gene expression. The left figure shows that the SNP influences methylation levels that in turn influence gene expression levels. The right figure shows the directed acyclic graph that represents this model. Faded symbols represent the measured values whereas solid symbols represent the true values. b) The same as in A, except the causal direction is from gene expression to DNA methylation. c) A model of confounding, where gene expression and DNA methylation are not causally related, but the SNP influences them each through separate pathways or a confounder.",
	display=FALSE
)

fig_nums(
	"cit_measurement_error_figure", 
	caption="The CIT was performed on simulated variables where the exposure influenced the outcome and the exposure was instrumented by a SNP. The test statistic from CIT when testing if the exposure caused the outcome (the true model) is in red, and the test for the outcome causing the exposure (false model) is in green. Rows of plots represent the sample sizes used for the simulations. As measurement error increases (decreasing values on x-axis) the test statistic for the incorrect model gets stronger and the test statistic for the correct model gets weaker.",
	display=FALSE
)

@@

fig_nums(
	"shakhplot",
	caption=paste0("Using ", nrow(qtl_orig), " putative associations between DNA methylation and gene expression we used the Steiger test to infer the direction of causality between them. a) The rightmost bar shows the proportion of associations for each of the two possible causal directions (colour key) assuming no measurement error in either gene expression or DNA methylation levels. The proportions change when we assume different levels of measurement error in gene expression levels (x-axis) or DNA methylation levels (columns of boxes). If there is systematically higher measurement error in one platform it will appear to be less likely to be the causal factor. b) The relationship between the Pearson correlation between DNA methylation and gene expression levels (x-axis) and the causal estimate (scaled to be in standard deviation units, y-axis). c) Distribution of estimated causal effect sizes, stratified into associations inferred to be due to DNA methylation causing expression (blue) and expression causing DNA methylation (red)."),
	display=FALSE
)

@@

Inference of the causal structure that induces correlations between two traits can be achieved by combining genetic associations with a mediation-based approach, as is done in the causal inference test (CIT) and others. However, we show that measurement error in the phenotypes can lead to mediation-based approaches inferring the wrong causal direction, and that increasing sample sizes has the adverse effect of increasing confidence in the wrong answer. Here we introduce an extension to Mendelian randomisation, a method that uses genetic associations in an instrumentation framework, that enables inference of the causal direction with some advantages. First, it is less susceptible to bias in the presence of measurement error; second, it is more statistically efficient; third, it can be performed using only summary level data from genome-wide association studies; and fourth, its sensitivity to measurement error can be evaluated. We apply the method to infer the causal direction between DNA methylation and gene expression levels. Our results demonstrate that, in general, DNA methylation is more likely to be the causal factor, but that such high level conclusions are susceptible to differential measurement error between the platforms. We emphasise that the use of several different methods is important to triangulate reliable conclusions about causality.

@@

Measurement error (or observational error, henceforth referred to simply as measurement error) is the difference between the measured value of a quantity and its true value. This study focuses specifically on non-differential measurement error where all strata of a measured variable have the same error rate, which can manifest as changes in scale or levels of imprecision (noise) of the measurement. Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009]. Array technology is now commonly used to obtain high throughput phenotyping at low cost, but comes with the problem of having imperfect resolution, for instance methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value due to imperfect sensitivity [@Harper2013; @Chen2013a]. Relatedly, if the measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].

Measurement error will of course arise in more basic data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately. Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the epidemiological literature extensively [@LeCessie2012; @Blakely2013]. Given the near-ubiquitous presence of measurement error in phenomic data it is vital to understand its impact on the tools we use for causal inference.

@@

By utilising genetic instruments in different ways, mediation-based analysis and MR models have properties that confer some advantages and some disadvantages for reliable causal inference. An important advantage of mediation-based analysis is that the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure. In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence. The CIT can also distinguish between several putative models of association between the traits and the SNP [@Millstein2009; @Millstein2016]. Such is not the case for MR, where in order to infer the direction of causality between two traits the instrument must be known to influence the exposure primarily, associating with the outcome only through the exposure.

Assuming biological knowledge of genetic associations can be problematic because if there exists a putative association between two variables, with the instrumental SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the primary effect of the SNP (i.e. for which of the two variables is the SNP a valid instrument? `r cfig("dag_figure")`). By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, such that if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure. Such a situation can arise in many scenarios. Genome-wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015]. If multiple instruments are available for an hypothesised exposure, which is increasingly typical for complex traits that are analysed in large GWAS consortia, then techniques can be applied to mitigate these issues [@DaveySmithHemani2014]. But these techniques cannot always be applied in the case of determining causal directions between 'omic measures where typically only one cis-acting SNP is known. For example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level [@Shakhbazov2016] (`r cfig("dag_figure")`)?

@@

Using theory and simulations we show how non-differential measurement error in phenotypes can lead to unreliable causal inference in the mediation-based CIT method. We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a method to evaluate the sensitivity of the result of this extension to measurement error. This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially. Finally, we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels.

@@

The Steiger test for inferring if $x \rightarrow y$ is based on evaluating $\rho_{gx} > \rho_{gy}$. However, $\rho_{gx}$ and $\rho_{gy}$ are underestimated if $x$ and $y$ are measured imprecisely. If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\rho_{g,x_o} > \rho_{g,y_o}$. As we show in the appendix, the maximum measurement error of $x_o$ is $\rho_{g,x_o}$ and the minimum is 0 (same applies to $y_o$). It is possible to simulate what the inferred causal direction would be for all values within these bounds. To evaluate how reliable, $R$, the Steiger test is to potential measurement imprecision in $x$ and $y$ we integrate over the simulated $R$ values for the possible measurement error values which are compatible with the direction of causality obtained empirically from the Steiger test. We also do the same for the measurement error values which are not compatible. The ratio of these quantities is an indication of how much of the measurement error space supports the inferred causal direction. A ratio $R=1$ indicates that the Steiger test is highly sensitive to measurement error, with equal weight of the measurement error parameter space supporting both directions of causality. Full details are provided in Appendix 2.

@@

Two sample MR [@Pierce2013] was performed using the summary statistics presented in [@Shakhbazov2016] for each of `r nrow(shakhbazov)` associations between DNA methylation and gene expression levels. To do this we obtained a list of `r nrow(qtl_orig)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016]. These were filtered to be located on the same chromosome, have robost correlations after correcting for multiple testing, and to share a SNP that had a robust cis-acting effect on both the DNA methylation probe and the gene expression probe. Because only summary statistics were available (effect, standard error, effect allele, sample size, p-values) for the instrumental SNP on the methylation and gene expression levels, the Steiger test of two independent correlations was used to infer the direction of causality for each of the associations. The Wald ratio test was then used to estimate the causal effect size for the estimated direction for each association.

@@

### MR analysis of gene expression and DNA methylation levels

We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels and found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a). We then went on to predict the causal directions of the associations for varying levels of possible measurement error. `r cfig("shakhplot")`a shows that the qualitative result of which of DNA methylation or gene expression is more commonly the causal factor is very strongly determined by measurement error.

We performed two sample MR [@Pierce2013] for each association in the direction of causality that we estimated to be the correct direction from the Steiger test. We observed that the sign of the MR estimate was generally in the same direction as the Pearson correlation coefficient reported by Shakhbazov et al [-@Shakhbazov2016], but with a few exceptions (`r cfig("shakhplot")`b). There was a moderate correlation between the absolute magnitudes of the causal correlation and the observational Pearson correlation (r = `r cor(abs(shakhbazov$mr_r[real_index]), abs(shakhbazov$pearson[real_index]))`). Together these inferences suggest that even for relatively low level associations it is important to use causal inference methods over observational associations to infer causal effect sizes.

We also observed that for associations where methylation caused gene expression the causal effect was more likely to be negative than for the associations where gene expression caused methylation (OR = `r shakhtest2$estimate` (95% CI `r shakhtest2$conf.int[1]` - `r shakhtest2$conf.int[2]`), `r cfig("shakhplot")`c), suggesting that reducing gene expression levels at a controlling CpG typically leads to increased gene expression levels, consistent with expectation [@Bird2002].

@@

Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data. In the epidemiological literature issues of measurement error in mediation analysis is relatively well explored [@Fewell2007]. Our analysis extends this to related methods such as CIT that are widely used in predominantly 'omic data. These methods are indeed liable to the same issues as standard mediation based analysis, and specifically we show that as measurement error in the exposure variable increases, CIT is likely to have reduced statistical power, and liable to infer the wrong direction of causality. We also demonstrate that, though unintuitive, increasing sample size does not resolve the issue, rather it leads to more extreme p-values for the model that predicts the wrong direction of causality.

Under many circumstances a practical solution to this problem is to use Mendelian randomisation instead of methods such as the CIT or similar that are based on mediation. Inferring the existence of causality using Mendelian randomisation is robust in the face of measurement error and, if the researcher has knowledge about the biology of the instrument being used in the analysis, can offer a direct solution to the issues that the CIT faces. This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance for the trait of interest. But on many occasions this is not the case, and it may be tempting to resort to using methods based on mediation in order to be able to both ascertain if there is a causal association and to infer the direction of causality. Here we have described a simple extension to MR which can be used as an alternative to mediation based methods. We show that this method is still liable to measurement error, but because it has different properties to the CIT it offers two main advantages. First, it uses a formal statistical framework to test for the robostness of the assumed direction of causality. Second, after testing in a comprehensive range of scenarios the MR based approach is less likely to infer the wrong direction of causality compared to CIT, while substantially improving power over CIT in the cases where $d > 0$.

@@

Mediation based network approaches, that go beyond analyses of two variables, are very well established [@Lagani2015] and have a number of extensions that make them valuable tools, including for example network construction. But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error. Recent advances in MR methodology, for example applying MR to genetical genomics [@Relton2012], multivariate MR [@Burgess2014a] and mediation through MR [@Varbo2015; @Burgess2015] may offer more robust alternatives for these more complicated problems.

>>>>

fig_nums(
	"dag_figure",
	caption="Gene expression levels (blue blocks) and DNA methylation levels (green triangles) may be correlated but the causal structure is unknown. If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are three basic competing models for these variables. The causal inference test (CIT) attempts to delineate between them. a) Methylation causes gene expression. The left figure shows that the SNP influences methylation levels that in turn influence gene expression levels. The right figure shows the directed acyclic graph that represents this model. Faded symbols represent the measured values whereas solid symbols represent the true values. b) The same as in A, except the causal direction is from gene expression to DNA methylation. c) A model of confounding, where gene expression and DNA methylation are not causally related, but the SNP influences them each through separate pathways or a confounder.",
	display=FALSE
)

fig_nums(
	"cit_measurement_error_figure", 
	caption="The CIT was performed on simulated variables where the exposure influenced the outcome and the exposure was instrumented by a SNP. The test statistic from CIT when testing if the exposure caused the outcome (the true model) is in red, and the test for the outcome causing the exposure (false model) is in green. Rows of plots represent the sample sizes used for the simulations. As measurement imprecision increases (decreasing values on x-axis) the test statistic for the incorrect model gets stronger and the test statistic for the correct model gets weaker.",
	display=FALSE
)

@@

fig_nums(
	"shakhplot",
	caption=paste0("Using ", nrow(qtl_orig), " putative associations between DNA methylation and gene expression we used the Steiger test to infer the direction of causality between them. a) The rightmost bar shows the proportion of associations for each of the two possible causal directions (colour key) assuming no measurement error in either gene expression or DNA methylation levels. The proportions change when we assume different levels of measurement error in gene expression levels (x-axis) or DNA methylation levels (columns of boxes). If there is systematically higher measurement error in one platform than the other it will appear to be less likely to be the causal factor. b) The relationship between the Pearson correlation between DNA methylation and gene expression levels (x-axis) and the causal estimate (scaled to be in standard deviation units, y-axis). c) Distribution of estimated causal effect sizes, stratified into associations inferred to be due to DNA methylation causing expression (blue) and expression causing DNA methylation (red)."),
	display=FALSE
)

@@

Inference of the causal structure that induces correlations between two traits can be achieved by combining genetic associations with a mediation-based approach, as is done in the causal inference test (CIT) and others. However, we show that measurement error in the phenotypes can lead to mediation-based approaches inferring the wrong causal direction, and that increasing sample sizes has the adverse effect of increasing confidence in the wrong answer. Here we introduce an extension to Mendelian randomisation, a method that uses genetic associations in an instrumentation framework, that enables inference of the causal direction between traits, with some advantages. First, it is less susceptible to bias in the presence of measurement error; second, it is more statistically efficient; third, it can be performed using only summary level data from genome-wide association studies; and fourth, its sensitivity to measurement error can be evaluated. We apply the method to infer the causal direction between DNA methylation and gene expression levels. Our results demonstrate that, in general, DNA methylation is more likely to be the causal factor, but this result is highly susceptible to bias induced by systematic differences in measurement error between the platforms. We emphasise that, where possible, implementing MR alongside other approaches such as CIT is important to triangulate reliable conclusions about causality.

@@

Measurement error is the difference between the measured value of a quantity and its true value. This study focuses specifically on non-differential measurement error where all strata of a measured variable have the same error rate, which can manifest as changes in scale or levels of imprecision (noise) of the measurement. Such variability can arise through a whole plethora of mechanisms, which are often unique to the study design and difficult to avoid [@Houle2011; @Hernan2009]. Array technology is now commonly used to obtain high throughput phenotyping at low cost, but comes with the problem of having imperfect resolution, for instance methylation levels as measured by the Illumina450k chip are prone to have some amount of noise around the true value due to imperfect sensitivity [@Harper2013; @Chen2013a]. Relatedly, if the measurement of biological interest is the methylation level in a T cell, then measurement error of this value can be introduced by using methylation levels from whole blood samples because the measured value will be an assay of many cell types [@Houseman2012].

Measurement error will of course arise in other types of data too, for example when measuring BMI one is typically interested in using this as a proxy for adiposity, but it is clear that the correlation between BMI and underlying adiposity is not perfect [@Ahima2013]. A similar problem of biological misspecification is unavoidable in disease diagnosis, and measuring behaviour such as smoking or diet is notoriously difficult to do accurately. Measurement error can also be introduced after the data has been collected, for example the transformation of non-normal data for the purpose of statistical analysis will lead to a new variable that will typically incur both changes in scale and imprecision (noise) compared to the original variable. The sources of measurement error are not limited to this list [@Hernan2009], and its impact has been explored in the epidemiological literature extensively [@LeCessie2012; @Blakely2013]. Given the near-ubiquitous presence of measurement error in phenomic data it is vital to understand its impact on the tools we use for causal inference.

@@

By utilising genetic instruments in different ways, mediation-based analysis and MR models have properties that confer some advantages and some disadvantages for reliable causal inference. An important advantage of mediation-based analysis is that in order to infer the causal direction the true underlying biology of the genetic instrument doesn't necessarily need to be known as long as it associates with the putative exposure. In the CIT framework (described fully in the Methods) for example, the test statistic is different if you test for the exposure causing the outcome or the outcome causing the exposure, allowing the researcher to infer the direction of causality between two variables by performing the test in both directions and choosing the model with the strongest evidence. The CIT also has the valuable property of being able to distinguish between several putative causal graphs that link the traits with the SNP (`r cfig("dag_figure")`). Such is not the case for MR, where in order to infer the direction of causality between two traits the instrument must be known to influence the exposure primarily, associating with the outcome only through the exposure.

Assuming biological knowledge of genetic associations can be problematic because if there exists a putative association between two variables, with the SNP being robustly associated with each, it can be difficult to determine which of the two variables is subject to the primary effect of the SNP (i.e. for which of the two variables is the SNP a valid instrument? `r cfig("dag_figure")`). By definition, we expect that if the association is causal then a SNP for the exposure will be associated with the outcome, such that if the researcher erroneously uses the SNP as an instrument for the outcome then they are likely to see an apparently robust causal association of outcome on exposure. Such a situation can arise in many scenarios. Genome-wide association studies (GWASs) that identify genetic associations for complex traits are, by design, hypothesis free and agnostic of genomic function, and it often takes years of follow up studies to understand the biological nature of a putative GWAS hit [@Claussnitzer2015]. If multiple instruments are available for an hypothesised exposure, which is increasingly typical for complex traits that are analysed in large GWAS consortia, then techniques can be applied to mitigate these issues [@DaveySmithHemani2014]. But these techniques cannot always be applied in the case of determining causal directions between 'omic measures where typically only one cis-acting SNP is known. For example if a DNA methylation probe is associated with expression of an adjacent gene, then is a cis-acting SNP an instrument for the DNA methylation level, or the gene expression level (`r cfig("dag_figure")`)?

@@

Using theory and simulations we show how non-differential measurement error in phenotypes can lead to unreliable causal inference in the mediation-based CIT method. We then present an extension to MR that allows researchers to ascertain the causal direction of an association even when the biology of the instruments are not fully understood, and also a method to evaluate the sensitivity of the result of this extension to measurement error. This improves the utility of MR in cases where mediation based methods might have otherwise been used preferentially. Finally, we apply this new method to infer the direction of causation between DNA methylation levels and gene expression levels. Our analyses highlight that because these different causal inference techniques have varying strengths and weaknesses, triangulation of evidence from as many sources as possible should be practiced in causal inference [@Lawlor2017].

@@

The Steiger test for inferring if $x \rightarrow y$ is based on evaluating $\rho_{gx} > \rho_{gy}$. However, $\rho_{gx}$ and $\rho_{gy}$ are underestimated if $x$ and $y$ are measured imprecisely. If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\rho_{g,x_o} > \rho_{g,y_o}$. As we show in the appendix, the maximum measurement error of $x_o$ is $\rho_{g,x_o}$ and the minimum is 0 (same applies to $y_o$). It is possible to simulate what the inferred causal direction would be for all values within these bounds. To evaluate how reliable, $R$, the Steiger test is to potential measurement imprecision in $x$ and $y$ we integrate over the simulated $R$ values for the possible measurement error values which are compatible with the direction of causality obtained empirically from the Steiger test. We also do the same for the measurement error values which are not compatible. The ratio of these quantities is an indication of how much of the measurement error space supports the inferred causal direction. A ratio $R=1$ indicates that the Steiger test is highly sensitive to measurement error, with equal weight of the measurement error parameter space supporting both directions of causality. Full details are provided in Appendix 2 and `r cfig("steiger_sensitivity_plot")`a.

@@

Two sample MR [@Pierce2013] was performed using the summary statistics presented in [@Shakhbazov2016] for each of `r nrow(shakhbazov)` associations between DNA methylation and gene expression levels. To do this we obtained a list of `r nrow(qtl_orig)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016]. These were filtered to be located on the same chromosome, have robust correlations after correcting for multiple testing, and to share a SNP that had a robust cis-acting effect on both the DNA methylation probe and the gene expression probe. Because only summary statistics were available (effect, standard error, effect allele, sample size, p-values) for the instrumental SNP on the methylation and gene expression levels, the Steiger test of two independent correlations was used to infer the direction of causality for each of the associations. The Wald ratio test was then used to estimate the causal effect size for the estimated direction for each association.

@@

### The causal relationship between gene expression and DNA methylation levels

We used the Steiger test to infer the direction of causality between DNA methylation levels and gene expression levels and found that the causal direction commonly goes in both directions (`r cfig("shakhplot")`a), but assuming no or equal measurement error DNA methylation levels were the predominant causal factor ($p = `r signif(shakhtest4$p.value, 2)`$). We then went on to predict the causal directions of the associations for varying levels of possible measurement error. `r cfig("shakhplot")`a shows that the qualitative result of which of DNA methylation or gene expression is more commonly the causal factor is very strongly determined by measurement error.

We performed two sample MR [@Pierce2013] for each association in the direction of causality that we estimated to be the correct direction from the Steiger test. We observed that the sign of the MR estimate was generally in the same direction as the Pearson correlation coefficient reported by Shakhbazov et al [-@Shakhbazov2016]  (`r cfig("shakhplot")`b). There was a moderate correlation between the absolute magnitudes of the causal correlation and the observational Pearson correlation (r = `r round(cor(abs(shakhbazov$mr_r[real_index]), abs(shakhbazov$pearson[real_index])), 2)`). Together these inferences suggest that even for relatively low level associations it is important to use causal inference methods over observational associations to infer causal effect sizes.

We also observed that for associations where methylation caused gene expression the causal effect was more likely to be negative than for the associations where gene expression caused methylation (OR = `r round(shakhtest2$estimate,2)` (95% CI `r round(shakhtest2$conf.int[1],2)` - `r round(shakhtest2$conf.int[2],2)`), `r cfig("shakhplot")`c), suggesting that reducing gene expression levels at a controlling CpG typically leads to increased gene expression levels, consistent with expectation [@Bird2002].

@@

Researchers are often confronted with the problem of making causal inferences using a statistical framework on observational data. In the epidemiological literature issues of measurement error in mediation analysis is relatively well explored [@Cole2014]. Our analysis extends this to related methods such as CIT that are widely used in predominantly 'omic data. These methods are indeed liable to the same issues as standard mediation based analysis, and specifically we show that as measurement error in the exposure variable increases, CIT is likely to have reduced statistical power, and liable to infer the wrong direction of causality. We also demonstrate that, though unintuitive, increasing sample size does not resolve the issue, rather it leads to more extreme p-values for the model that predicts the wrong direction of causality.

Under many circumstances a practical solution to this problem is to use Mendelian randomisation instead of methods such as the CIT or similar that are based on mediation. Inferring the existence of causality using Mendelian randomisation is robust in the face of measurement error and, if the researcher has knowledge about the biology of the instrument being used in the analysis, can offer a direct solution to the issues that the CIT faces. This assumption is often reasonable, for example SNPs are commonly used as instruments when they are found in genes with known biological relevance for the trait of interest. But on many occasions, especially in the realm of 'omic data, this is not the case, and methods based on mediation in have been valuable in order to be able to both ascertain if there is a causal association and to infer the direction of causality. Here we have described a simple extension to MR which can be used as an alternative to or in conjunction with mediation based methods. We show that this method is still liable to measurement error, but because it has different properties to the CIT it offers several main advantages. First, it uses a formal statistical framework to test for the robustness of the assumed direction of causality. Second, after testing in a comprehensive range of scenarios the MR based approach is less likely to infer the wrong direction of causality compared to CIT, while substantially improving power over CIT in the cases where $d > 0$.

@@

Mediation based network approaches, that go beyond analyses of two variables, are very well established [@Lagani2015] and have a number of extensions that make them valuable tools, including for example network construction. But because they are predicated on the basic underlying principles of mediation they are liable to suffer from the same issues of measurement error. Recent advances in MR methodology, for example applying MR to genetical genomics [@Relton2012], multivariate MR [@Burgess2014a] and mediation through MR [@Varbo2015; @Burgess2015; @Richmond2016] may offer more robust alternatives for these more complicated problems.
